---
title: "Holidays"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: "hide"
---

```{r setup, include = FALSE}
library(tidyverse)
library(patchwork)
library(plotly)
library(leaflet)
library(lubridate)
knitr::opts_chunk$set(
echo = TRUE,
warning = FALSE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%")
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis")
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

```{r basic data framing and tidying}
nypd_shooting_df = 
  read_csv("data/nypd_shooting_data.csv") %>%
  janitor::clean_names() %>%
  separate(col = occur_date, into = c("month", "day", "year"), sep = "/") %>% 
  separate(col = occur_time, into = c("hour", "minute", "second"), sep = ":") %>%
  mutate(across(where(is.character), tolower),
         month = as.numeric(month),
         month_name = recode(month, "1" = "january", "2" = "february", "3" = "march", "4" = "april", "5" = "may", "6" = "june", "7" = "july", "8" = "august", "9" = "september", "10" = "october", "11" = "november", "12" = "december"),
         day = as.numeric(day),
         year = as.numeric(year), 
         hour = as.numeric(hour),
         minute = as.numeric(minute), 
         second = as.numeric(second),
         minute_calc = hour * 60 + minute,
         boro = as.factor(boro), 
         boro = fct_relevel(boro, "manhattan", "brooklyn", "bronx", "queens", "staten island"),
         date = make_date(year = year, month = month, day = day)) %>% 
  relocate(incident_key, date)
```

# Background

We were interested to see if shooting incidents occurred in different frequency on New York City holidays compared to other days of the year. We hypothesized that both shooting victims and perpetrators would be more likely to be juveniles on holidays, as these days would represent periods in which school was not in session and therefore potential opportunities for social disruption.


# Methods

We defined NYC holidays as specific calendar days designed as state holidays in New York State. On some instances a designated holiday was assigned as the day upon which the holiday was _observed_ rather than on the typical/historic day of observation. For example, New Year's Day is a state-observed holiday in New York State. January 1, 2012 occurred on Sunday (a typical weekend/non-working day). Thus, Monday, January 2, 2012 was designed as New Year's Day in accordance with state observation and was labeled in our analysis as a holiday.

We included the the following New York State-recognized holidays: New Year's Day, Martin Luther King Jr. Day, Lincoln's Birthday, Washington's Birthday, Memorial Day, Juneteenth, Independence Day, Labor Day, Columbus Day, Veteran's Day, Thanksgiving, and Christmas.

Lincoln's Birthday and Washington's Birthday are designated as individual holidays in New York State. These typically occur on February 12th and 15th, respectively. In other states, these days are observed in combination on the third Monday in February. In some analyses, we combined Lincoln's and Washington's Birthdays as one holiday to allow for equitable comparison of shooting incidence and demographics to other holidays.

We wrote a function to identify these dates individually within the publicly-available dataset:
```{r function for holidays 2010 - 2021}
holidays = function(x) {
  if (x == "2010-01-01" | x == "2012-01-02" | x == "2013-01-01" | x ==	"2014-01-01" | x ==	"2015-01-01"| x == "2016-01-01" | x == "2017-01-02" | x ==	"2018-01-01" | x == "2019-01-01" | x ==	"2020-01-01" | x ==	"2021-01-01") {"new year's day observed"}
  else if (x == "2010-01-18"| x == "2011-01-17"| x ==	"2012-01-16"| x ==	"2013-01-21"| x ==	"2014-01-20"| x ==	"2015-01-19"| x ==	"2016-01-18"| x ==	"2017-01-16"| x ==	"2018-01-15"| x ==	"2019-01-21"| x ==	"2020-01-20"| x ==	"2021-01-18") {"mlk day"}
  else if (x == "2010-02-12"| x ==	"2011-02-12"| x ==	"2012-02-13"| x ==	"2013-02-12"| x ==	"2014-02-12"| x ==	"2015-02-12"| x ==	"2016-02-12"| x ==	"2017-02-13"| x ==	"2018-02-12"| x ==	"2019-02-12"| x ==	"2020-02-12"| x ==	"2021-02-12") {"lincoln's bday"}
  else if (x == "2010-02-15"| x ==	"2011-02-21"| x ==	"2012-02-20"| x ==	"2013-02-18"| x ==	"2014-02-17"| x ==	"2015-02-16"| x ==	"2016-02-15"| x ==	"2017-02-20"| x ==	"2018-02-19"| x ==	"2019-02-18"| x ==	"2020-02-17"| x ==	"2021-02-15") {"washington's bday"}
  else if (x == "2010-05-31"| x ==	"2011-05-30"| x ==	"2012-05-28"| x ==	"2013-05-27"| x ==	"2014-05-26"| x ==	"2015-05-25"| x ==	"2016-05-30"| x ==	"2017-05-29"| x ==	"2018-05-28"| x ==	"2019-05-27"| x ==	"2020-05-25"| x ==	"2021-05-31") {"memorial day"}
    else if (x == "2010-06-19"| x == "2011-06-19" | x == "2012-06-19" | x == "2013-06-19" | x == "2014-06-19" | x ==	"2015-06-19" | x ==	"2016-06-19" | x ==	"2017-06-19" | x ==	"2018-06-19" | x ==	"2019-06-19" | x ==	"2020-06-19" | x ==	"2021-06-19") {"juneteenth"}
  else if (x == "2010-07-04"| x ==	"2011-07-04"| x ==	"2012-07-04"| x ==	"2013-07-04"| x ==	"2014-07-04"| x ==	"2015-07-03"| x ==	"2016-07-04"| x ==	"2017-07-04"| x ==	"2018-07-04"| x ==	"2019-07-04"| x ==	"2020-07-03"| x ==	"2021-07-05") {"independence day"}
  else if (x == "2010-09-06"| x ==	"2011-09-05"| x ==	"2012-09-03"| x ==	"2013-09-02"| x ==	"2014-09-01"| x ==	"2015-09-07"| x ==	"2016-09-05"| x ==	"2017-09-04"| x ==	"2018-09-03"| x ==	"2019-09-02"| x ==	"2020-09-07"| x ==	"2021-09-06") {"labor day"}
  else if (x == "2010-10-11"| x ==	"2011-10-11"| x ==	"2012-10-08" | x ==	"2013-10-14"| x ==	"2014-10-13"| x ==	"2015-10-12"| x ==	"2016-10-10"| x ==	"2017-10-09"| x ==	"2018-10-08"| x ==	"2019-10-14"| x ==	"2020-10-12"| x ==	"2021-10-11") {"columbus day"}
  else if (x == "2010-11-11"| x ==	"2011-11-11"| x ==	"2012-11-12"| x ==	"2013-11-11"| x ==	"2014-11-11"| x ==	"2015-11-11"| x ==	"2016-11-11"| x ==	"2017-11-10"| x ==	"2018-11-12"| x ==	"2019-11-11"| x ==	"2020-11-11"| x ==	"2021-11-11") {"veteran's day"}
  else if (x == "2010-11-25"| x ==	"2011-11-24"| x ==	"2012-11-22"| x ==	"2013-11-28"| x ==	"2014-11-27"| x ==	"2015-11-26"| x ==	"2016-11-24"| x ==	"2017-11-23"| x ==	"2018-11-22"| x ==	"2019-11-28"| x ==	"2020-11-26"| x ==	"2021-11-25") {"thanksgiving"}
  else if (x == "2010-12-24"| x ==	"2011-12-26"| x ==	"2012-12-25"| x ==	"2013-12-15"| x ==	"2014-12-25"| x ==	"2015-12-25"| x ==	"2016-12-26"| x ==	"2017-12-25"| x ==	"2018-12-25"| x ==	"2019-11-25"| x ==	"2020-12-25"| x ==	"2021-12-24") {"christmas observed"}
  else {"non-holiday"}
}
```

We chose to limit our analyses to years **2010 - 2021**, as this would represent a full decade of data, and applied our holiday function to the data.
```{r nypd shootings on holidays dataframe}
# Limiting data to years 2010 - 2021
# This takes several minutes to run due to mapping many criteria to a large dataset
nypd_shooting_holidays = 
  nypd_shooting_df %>% 
  filter(
    year >= 2010) %>% 
  mutate(
    holiday_summary = map(date,holidays)) %>% 
  relocate(incident_key, date, holiday_summary) %>% 
  unnest(holiday_summary) 
```

# Results


### Yearly Variation in Holiday Shootings

We explored year-to-year variation in shootings in NYC, by holiday.

```{r bar plot count per year in each boro by holiday}
nypd_shooting_holidays %>% 
  filter(
    holiday_summary != "non-holiday")%>% 
  group_by(year, boro, holiday_summary) %>% 
  summarize(
    count = n()) %>% 
  plot_ly(x = ~year, y = ~count, color = ~holiday_summary, type = "bar") %>% 
  layout(yaxis = list(title = "Count"), barmode = "stack") %>% 
  layout(xaxis = list(title = "Year"))
```

2021 has a greater number of shootings on holidays than prior years, though 2010 and 2011 also had a large number of holiday shootouts.


### Safe vs Lethal Holidays

We were interested to see if some holidays are safer than other holidays. (We combined Lincoln's and Washington's birthdays into one "President's Day" holiday for this analysis.)

```{r total shootings 2010 - 2021 per holiday}
nypd_shooting_holidays %>% 
  filter(
    holiday_summary != "non-holiday") %>% 
  # Lincoln's Birthday and President's Birthday combined as "President's Day"
  mutate(
    holiday_summary = str_replace(holiday_summary, "washington's bday","president's day"),
    holiday_summary = str_replace(holiday_summary, "lincoln's bday","president's day")) %>% 
  count(holiday_summary) %>% 
  mutate(
    holiday_summary = fct_reorder(holiday_summary,n)) %>% 
  plot_ly(x = ~holiday_summary, y = ~n, color = ~holiday_summary, type = "bar") %>% 
  layout(yaxis = list(title = "Count"), xaxis = list(title = "Holiday"), title = "NYC Shootings Per Holiday, 2010 - 2021") %>% 
  hide_legend()
```

**In general, holidays that occur in summer months have a larger number of shootings than those in the fall or winter.**


### Perpetrator Age Distribution on Non-holidays compared to Holidays

We explored whether the range of shooters' ages appeared subjectively different on non-holidays compared to holidays.

```{r perp ages histograms}
perp_ages_non_holidays_histogram = 
nypd_shooting_holidays %>% 
  filter(
    holiday_summary == "non-holiday"
  ) %>% 
  # omitting age ranges of 1020, 224, and 940 which have 1 incident each and may be typos
  mutate(
    perp_age_group = str_replace(perp_age_group, "1020", "typo"),
    perp_age_group = str_replace(perp_age_group, "224","typo"),
    perp_age_group = str_replace(perp_age_group, "940","typo")) %>% 
  drop_na(perp_age_group) %>% 
  filter(perp_age_group != "typo") %>% 
  plot_ly(x = ~perp_age_group, type = "histogram") %>% 
  layout(xaxis = list(title = "Perpetrator Age Group"), yaxis = list(title = "Count"))

perp_age_holidays_histogram = 
nypd_shooting_holidays %>% 
  filter(
    holiday_summary != "non-holiday"
  ) %>% 
  drop_na(perp_age_group) %>% 
  plot_ly(x = ~perp_age_group, type = "histogram") %>% 
  layout(xaxis = list(title = "Perpetrator Age Group"), yaxis = list(title = "Count"))

subplot(perp_ages_non_holidays_histogram, perp_age_holidays_histogram, shareX = TRUE, shareY = TRUE, titleX = FALSE, titleY = TRUE)
```

In this plot, `trace 0` represents non-holidays and `trace 1` represents holidays. Clearly, shootings on non-holidays are more common than on holidays. However, it is unclear in this histogram if the proportion of juvenile perpetrators is different on holidays than on non-holidays.


### Victim  Age Distribution on Non--holidays compared to Holidays

Similarly, we explored whether the range of victims' ages appeared subjectively different on non-holidays compared to holidays.

```{r victim ages histograms}
vic_ages_non_holidays_histogram = 
nypd_shooting_holidays %>% 
  filter(
    holiday_summary == "non-holiday"
  ) %>% 
  drop_na(vic_age_group) %>% 
  plot_ly(x = ~vic_age_group, type = "histogram") %>% 
  layout(xaxis = list(title = "Victim Age Group"), yaxis = list(title = "Count"))


vic_ages_holidays_histogram = 
nypd_shooting_holidays %>% 
  filter(
    holiday_summary != "non-holiday"
  ) %>% 
  drop_na(vic_age_group) %>% 
  plot_ly(x = ~vic_age_group, type = "histogram") %>% 
  layout(xaxis = list(title = "Victim Age Group"), yaxis = list(title = "Count"))

subplot(vic_ages_non_holidays_histogram, vic_ages_holidays_histogram, shareX = TRUE, shareY = TRUE, titleX = FALSE, titleY = TRUE)
```

In this plot, `trace 0` represents non-holidays and `trace 1` represents holidays. Clearly, shootings on non-holidays are more common than on holidays. However, it is unclear in this histogram if the proportion of juvenile victims is different on holidays than on non-holidays.

### Is there a significant difference in the proportion of shooting perpetrators that are < 18 years old compared to perpetrators > 18 years old on holidays compared to non-holidays?

As above, we hypothesized that the proportion of juvenile _perpetrators_ (< 18 years old) compared to adult perpetrators (> 18 years old) would be different on holidays compared to non-holidays. We performed a two-sample test of proportions to test this theory (versus the null hypothesis that juveniles would perform equivalent proportion of shootings on holidays compared to non-holidays). Statistical significance was defined as p < 0.05

```{r prop test perp age <18 vs adult on holiday vs non-holiday}
holiday_binary_perp_age_df = 
nypd_shooting_holidays %>% 
    # omitting age ranges of 1020, 224, and 940 which have 1 incident each and may be typos
  mutate(
    perp_age_group = str_replace(perp_age_group, "1020", "typo"),
    perp_age_group = str_replace(perp_age_group, "224","typo"),
    perp_age_group = str_replace(perp_age_group, "940","typo")) %>% 
  drop_na(vic_age_group) %>% 
  drop_na(perp_age_group) %>% 
  filter(perp_age_group != "typo") %>% 
  mutate(
    holiday_binary = ifelse(holiday_summary != "non-holiday","holiday","non-holiday"),
    perp_juvenile = ifelse(perp_age_group == "<18","juvenile","adult")) %>% 
  group_by(perp_juvenile, holiday_binary) %>% 
  summarize(
    count = n()) %>%
  pivot_wider(
    names_from = perp_juvenile,
    values_from = count) %>%
  janitor::clean_names() %>% 
  relocate(holiday_binary, juvenile) %>% 
  mutate(
    total = juvenile + adult,
    proportion_juvenile = juvenile/total)

holiday_binary_perp_age_df %>% 
  knitr::kable(digits = 3)

prop.test(holiday_binary_perp_age_df$juvenile,holiday_binary_perp_age_df$total) %>% 
  broom::tidy() %>% 
  select(-estimate1, -estimate2, -statistic, -parameter, -alternative) %>%
  knitr::kable(digits = 3)
```

**We did not identify a significant difference in the proportion of shootings perpetrated by juveniles compared to non-juveniles on holidays versus non-holidays.**


### Is there a significant difference in the proportion of shooting victims who are < 18 years old compared to victims > 18 years old on holidays compared to non-holidays?

Finally, we hypothesized that the proportion of juvenile _victims_ (< 18 years old) compared to adult victims (> 18 years old) would be significantly different on holidays compared to non-holidays. We performed a two-sample test of proportions to test this theory (versus the null hypothesis that proportion of juvenile shooting victims would be equivalent on holidays compared to non-holidays). Statistical significance was defined as p < 0.05

```{r prop test victim age <18 vs adult on holiday vs non-holiday}
holiday_binary_vic_age_df = 
nypd_shooting_holidays %>% 
  drop_na(vic_age_group) %>% 
  mutate(
    holiday_binary = ifelse(holiday_summary != "non-holiday","holiday","non-holiday"),
    vic_juvenile = ifelse(vic_age_group == "<18","juvenile","adult")) %>% 
  group_by(vic_juvenile, holiday_binary) %>% 
  summarize(
    count = n()) %>%
  pivot_wider(
    names_from = vic_juvenile,
    values_from = count) %>%
  janitor::clean_names() %>% 
  relocate(holiday_binary, juvenile) %>% 
  mutate(
    total = juvenile + adult,
    proportion_juvenile = juvenile/total)

holiday_binary_vic_age_df %>% 
  knitr::kable(digits = 3)

prop.test(holiday_binary_vic_age_df$juvenile,holiday_binary_vic_age_df$total) %>% 
  broom::tidy() %>% 
  select(-estimate1, -estimate2, -statistic, -parameter, -alternative) %>%
  knitr::kable(digits  = 3)
```

**We did not identify a significant difference in the proportion of juvenile compared to non-juvenile victims of shootings on holidays versus non-holidays.**

# Conclusion

Holidays have fewer shooting incidents than non-holidays. Some holidays are "safer" than others, though this may represent seasonal variation rather than a distinct effect of the celebratory cause itself. Juveniles commit (and suffer) a comparable proportion of shootings to adults on holidays as they do on non-holidays.