---
title: "Overview"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: "hide"
---

```{r setup, include = FALSE}
library(tidyverse)
library(patchwork)
library(plotly)
library(leaflet)
library(gganimate)
library(gifski)


knitr::opts_chunk$set(
echo = TRUE,
warning = FALSE,
message = FALSE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%")

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis")

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

```{r}
nypd_shooting_df = 
  read_csv("data/nypd_shooting_data.csv") %>%
  janitor::clean_names() %>%
  separate(col = occur_date, into = c("month", "day", "year"), sep = "/") %>%
  separate(col = occur_time, into = c("hour", "minute", "second"), sep = ":") %>%
  mutate(across(where(is.character), tolower),
         month = as.numeric(month),
         month_name = recode(month, "1" = "january", "2" = "february", "3" = "march", "4" = "april", "5" = "may", "6" = "june", "7" = "july", "8" = "august", "9" = "september", "10" = "october", "11" = "november", "12" = "december"),
         day = as.numeric(day),
         year = as.numeric(year), 
         hour = as.numeric(hour),
         minute = as.numeric(minute), 
         second = as.numeric(second),
         minute_calc = hour * 60 + minute,
         boro = as.factor(boro), 
         boro = fct_relevel(boro, "manhattan", "brooklyn", "bronx", "queens", "staten island")) %>%
  select(incident_key, year, month_name, month, day, hour, minute, second, minute_calc, everything())
```

Here is a summary of the overall demographics of the perpetrators and victims involved in NYC shootings, including age, sex, and race/ethnicity. There were a total of `r nrow(nypd_shooting_df)` shootings from 2006 to 2021 reported by the NYPD. Overall, we see that <b>the majority of victims and perpetrators were young (aged 18-44 years), black, and males.</b><br>

<em>*Of note, these data included many NA entries particularly for perpetrators, so analyses were conducted on available data. For the purposes of simplifying race/ethnicity categorization, we collapsed white Hispanic and black Hispanic individuals into a single category titled “Hispanic”.</em><br><br>

### Perpetrator Demographics

#### Age
<em>48.8% NA</em>
```{r}
nypd_shooting_df %>%
  mutate(
    perp_age_group = na_if(perp_age_group, 1020),
    perp_age_group = na_if(perp_age_group, 224),
    perp_age_group = na_if(perp_age_group, 940),
    perp_age_group = na_if(perp_age_group, "unknown"),
  ) %>%
  drop_na(perp_age_group) %>%
  group_by(perp_age_group) %>%
  summarise(
    count = n()) %>%
  mutate(
    percentage = count / sum(count) * 100,
    percentage = round(percentage, digits = 1)) %>%
  knitr::kable()
```

#### Sex
<em>42.2% NA</em>
```{r}
nypd_shooting_df %>%
   mutate(
    perp_sex = na_if(perp_sex, "u")) %>%
  drop_na(perp_sex) %>%
  group_by(perp_sex) %>%
  summarise(
    count = n()) %>%
  mutate(
    percentage = count / sum(count) * 100,
    percentage = round(percentage, digits = 1)) %>%
  knitr::kable()
```

#### Race/Ethnicity
<em>43.5% NA</em>
```{r}
nypd_shooting_df %>%
  mutate(
    perp_race = recode(perp_race, "black hispanic" = "hispanic", "white hispanic" = "hispanic"), 
    perp_race = na_if(perp_race, "unknown")
  ) %>%
  drop_na(perp_race) %>%
  group_by(perp_race) %>%
  summarise(
    count = n()) %>%
  mutate(
    percentage = count / sum(count) * 100,
    percentage = round(percentage, digits = 1)) %>%
  knitr::kable()
```

### Victim Demographics

#### Age
<em>0.2% NA</em>
```{r}
nypd_shooting_df %>%
  group_by(vic_age_group) %>%
  mutate(
    vic_age_group = na_if(vic_age_group, "unknown")
  ) %>%
  summarise(
    count = n()) %>%
  mutate(
    percentage = count / sum(count) * 100,
    percentage = round(percentage, digits = 1)) %>%
  knitr::kable()
```

#### Sex
<em>0.04% NA</em>
```{r}
nypd_shooting_df %>%
  group_by(vic_sex) %>%
  mutate(
    vic_sex = na_if(vic_sex, "u")) %>%
  drop_na(vic_sex) %>%
  summarise(
    count = n()) %>%
  mutate(
    percentage = count / sum(count) * 100,
    percentage = round(percentage, digits = 1)) %>%
  knitr::kable()
```

#### Race/Ethnicity
<em>0.3% NA</em>
```{r}
nypd_shooting_df %>%
   mutate(
    vic_race = recode(vic_race, "black hispanic" = "hispanic", "white hispanic" = "hispanic"), 
    vic_race = na_if(vic_race, "unknown")
  ) %>%
  drop_na(vic_race) %>%
  group_by(vic_race) %>%
  summarise(
    count = n()) %>%
  mutate(
    percentage = count / sum(count) * 100,
    percentage = round(percentage, digits = 1)) %>%
  knitr::kable()
```

### Murders by Race/Ethnicity
We were also curious to see which race/ethnicity group suffered the highest mortality from shooting incidents from 2006 to 2021. The animation below shows that black individuals had the highest percentage of murders from shooting incidents overall, and by year, then Hispanics.

```{r}
scatter_df = 
  nypd_shooting_df%>%
  mutate(
    vic_race = recode(vic_race, "black hispanic" = "hispanic", "white hispanic" = "hispanic", "asian / pacific islander" = "asian_pacific"), 
  vic_race = na_if(vic_race, "unknown")) %>%
  filter (statistical_murder_flag == "TRUE", 
          vic_race != "NA") %>%
  relocate(vic_race, year) %>%
  group_by(year, vic_race) %>%
  count() %>%
  ungroup(year, vic_race)
  
relocate_race =
  nypd_shooting_df %>%
   relocate(vic_race, year) %>%
  mutate(
    vic_race = recode(vic_race, "black hispanic" = "hispanic", "white hispanic" = "hispanic", "asian / pacific islander" = "asian_pacific"), 
  vic_race = na_if(vic_race, "unknown")) %>%
  filter (statistical_murder_flag == "TRUE", 
          vic_race != "NA") 

anim_scatter =
  left_join(relocate_race, scatter_df, by = c("vic_race" = "vic_race", "year" = "year"))
 
by_year = 
  scatter_df %>%
  ungroup() %>%
  group_by(year) %>%
  pivot_wider(
    names_from = vic_race,
    values_from = n) %>%
  mutate(
    totalsbyyear = 
       as.numeric(asian_pacific) + as.numeric(black) +as.numeric(hispanic) + as.numeric(white)) %>%
pivot_longer(
  asian_pacific:white, 
  names_to = "vic_race",
  values_to = "n") %>%
  select (vic_race,year, n, totalsbyyear)
  
anim_scatter2 = 
  left_join(anim_scatter, by_year, by =c("year" = "year", "vic_race" = "vic_race","n" = "n")) 
  
anim_scatter_plot = 
  anim_scatter2 %>% 
  group_by(vic_race, year) %>% 
  mutate(
    percentage= (n/totalsbyyear * 100)) %>% 
  ggplot(aes(x = year, y = percentage, color = vic_race)) +
  geom_line() +
  geom_point() +
  labs(title = "Percentages of Shootings Resulting in Murders by Race/Ethnicity",
       x = "Year",
       y = "Percentage") +
  transition_reveal(year)
 
animate(anim_scatter_plot, renderer = gifski_renderer())
```

### "Raise the Age" Analysis
On April 10, 2017, New York State raised the age of criminal responsibility to 18 years of age, ensuring that young people in New York who commit non-violent crimes receive the intervention and evidence-based treatment they need. As of October 2019, New York no longer automatically prosecutes 16- and 17-year-olds as adults. This is known as the ["Raise the Age" law](https://ww2.nycourts.gov/ip/oji/raisetheage.shtml#:~:text=On%20April%2010%2C%202017%2C%20New,evidence%2Dbased%20treatment%20they%20need.). There has been concern that this new policy has resulted in a backlash of [increasing NYC teen shootings](
https://nypost.com/2022/09/25/raise-the-age-law-is-clearly-behind-soaring-nyc-teen-shootings/).

We sought to explore this trend and see if it holds true according to this dataset. <b>It is sobering to see that the number and percentage of shooters and victims aged <18 years in NYC has approximately doubled in the five years since "Raise the Age" in our analysis.</b>

```{r}
number_youthperp = 
  nypd_shooting_df %>%
  mutate(
    perp_age_group = na_if(perp_age_group, 1020),
    perp_age_group = na_if(perp_age_group, 224),
    perp_age_group = na_if(perp_age_group, 940),
    perp_age_group = na_if(perp_age_group, "unknown"),
  ) %>%
  drop_na(perp_age_group) %>%
  filter(year >= 2017, perp_age_group == "<18") %>%
  group_by(perp_age_group, year) %>%
  summarise(
    count = n()) %>%
  mutate(
    percentage = count / sum(count) * 100,
    percentage = round(percentage, digits = 1)) %>%
  ggplot(aes(x = year, y = count)) +
  geom_point() + geom_line() +
  labs(title = "# Shootings by <18 Year Olds") 

percent_youthperp = 
nypd_shooting_df %>%
  mutate(
    perp_age_group = na_if(perp_age_group, 1020),
    perp_age_group = na_if(perp_age_group, 224),
    perp_age_group = na_if(perp_age_group, 940),
    perp_age_group = na_if(perp_age_group, "unknown"),
  ) %>%
  drop_na(perp_age_group) %>%
  filter(year >= 2017, perp_age_group == "<18") %>%
  group_by(perp_age_group, year) %>%
  summarise(
    count = n()) %>%
  mutate(
    percentage = count / sum(count) * 100,
    percentage = round(percentage, digits = 1)) %>%
  ggplot(aes(x = year, y = percentage)) +
  geom_point() + geom_line() +
  labs(title = "% Shootings by <18 Year Olds") 

number_youthperp + percent_youthperp
```

```{r}
number_youthvic = 
  nypd_shooting_df %>%
  group_by(vic_age_group) %>%
  mutate(
    vic_age_group = na_if(vic_age_group, "unknown")
  ) %>%
  drop_na(vic_age_group) %>%
  filter(year >= 2017, vic_age_group == "<18") %>%
  group_by(vic_age_group, year) %>%
  summarise(
    count = n()) %>%
  mutate(
    percentage = count / sum(count) * 100,
    percentage = round(percentage, digits = 1)) %>%
  ggplot(aes(x = year, y = count)) +
  geom_point() + geom_line() +
  labs(title = "# Shooting Victims <18 Years") 

percent_vicyouth = 
  nypd_shooting_df %>%
  group_by(vic_age_group) %>%
  mutate(
    vic_age_group = na_if(vic_age_group, "unknown")
  ) %>%
  drop_na(vic_age_group) %>%
  filter(year >= 2017, vic_age_group == "<18") %>%
  group_by(vic_age_group, year) %>%
  summarise(
    count = n()) %>%
  mutate(
    percentage = count / sum(count) * 100,
    percentage = round(percentage, digits = 1)) %>%
  ggplot(aes(x = year, y = percentage)) +
  geom_point() + geom_line() +
  labs(title = "% Shooting Victims <18 Years") 

number_youthvic + percent_vicyouth
```
